Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-27 ThesisJP/Background/background.tex
Feedback from Jo Inge finished.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry in a recent essay \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation were also growing at the same rate until the required cooling finally became too much in 2004.  From that point, CPU manufactures had to pursue different directions than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders, which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and colors were fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other. Programmable shaders introduced the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this development was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.

The original graphics pipeline involved no conditional branching, and data were typically used once \cite{Purcell2002}. Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to get the most out of each processors. Balancing means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processor is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs. An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers. An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a high level of CPU-GPU data synchronization.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU. This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is designed to be the main control unit of a personal computer, and is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these features are and why there is still a big difference in theoretical peak performance between a CPU and a GPU.

@@

In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions. These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS}.
\end{equation}
Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores. A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)}).} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that the actual throughput is highly algorithm dependent. Nevertheless, in theory, the computational power provided by modern GPUs is five times higher than that provided by CPUs. A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}. However, in the next section we will see why it is often easier to obtain an optimized implementation for the GPU than for the CPU.

@@

\begin{figure}[t!]
\centering
\subfloat[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
}
\subfloat[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
}
\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an in-place element-wise matrix square.}
\label{fig:gpu_grid_kernel}
\end{figure}

@@

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction. The GPU is because of this sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor. This also explain why GPU threads are said to be light-weighted. Such a conceptual difference means that there is an inherent focus on breaking algorithms down into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Manually typed SIMT, or SIMD, instructions is the only way of programming the GPU, while a single thread with automatic vectorization of for-loops is the standard way of programming the CPU.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something that is not likely to happen in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is an incomplete and minimal example of a GPU program. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is part of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, programming the GPU will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction. The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, GPU programming will probably be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

@@

\subsection{Beamforming}
If time delays are applied to the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images that are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$ planes. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
\caption{Delay-and-sum array beamforming.}
\label{fig:das_background}
\end{figure}

@@

\subsection{Sampling and processing complexity}
When generating a B-mode image, we have to insure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system that both emits and receives energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

@@

To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field. This is done by estimating a sample covariance matrix, which for an active broadband system can be done in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] &= \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H\\
\vec{x}_l[n'] &= [x_l[n'], \dotso, x_{l+L}[n']]
\end{align}
where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a subarray of length $L$, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.  Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}. The number of temporal samples is selected relative to the pulse length. To ensure robustness against model errors and numerical stability the matrix is further loaded with a diagonal factor, 
\begin{align}
\epsilon &= d\times\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align} 
Hence, a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.

@@

\section{Volume rendering}\label{sec:volren}
\begin{figure}[t!]
\centering
\subfloat[Ray-casting]{
	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
}
\subfloat[Opacity transfer function]{
	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
}
\caption{The principle behind volume ray casting. a) A ray is casted out from every pixel in the output image. Along the ray, samples are gathered from the volume. The samples are assigned a color and an opacity before they are combined into one pixel value. b) The opacity transfer function used to determine how opaque a sample should be.}
\label{fig:vr}
\end{figure}
A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user. One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers. Another way is to project the whole volume into a two-dimensional image. The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image. Many different methods for volume rendering exist \cite{brodlie2001}, but in this thesis we will focus on ray casting \cite{Levoy1988}.

>>>>

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry in a recent essay \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation were also growing until the required cooling finally became too much in 2004.  From that point, CPU manufactures had to look at different strategies than just increasing the clock frequency. One strategy was to embed several cores in one CPU. This ensured continuation of Moore's law, since the number of transistors per chip (CPU) could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders in the field of graphics programming. The graphics processing pipeline had before programmable shaders been a fixed function pipeline, where geometry and colors were converted into shaded and rasterized sceneries by the graphics processing unit (GPU). Programmable shaders introduced the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this development was that GPUs were already available in consumer PCs, and hidden behind its graphics interface were processing capabilities an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.

The original graphics pipeline involved no conditional branching (caused by if-else control flow statements), and data were typically used once \cite{Purcell2002}. Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to get the most out of each processors. Balancing means that highly parallel and computationally intensive tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processor is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs. An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time can be saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize (or hide) CPU-to-GPU memory transfers. An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a lot of CPU-GPU data transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU. This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging systems.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is designed to be the main control unit of a personal computer, and is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these features are and why there is still a big difference in theoretical peak arithmetic performance between a CPU and a GPU.

@@

A special kind of instructions called SIMD was mentioned in Section \ref{sec:gpgpu}. These are instructions that are designed to execute a single instruction on multiple data elements in one clock cycle. In the late 1990s, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The recently introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply-add (\nom{FMA}{Fused multiply-add (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. The new extension AVX-512, that extends the registers to 512 bit, is found in the Intel accelerator board Xeon Phi, and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to a theoretical peak arithmetic performance of
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS},
\end{equation}
where FLOPS is short for; floating point operations per second. This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing capabilities to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia provides up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores. A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)} or SMX).} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) with close to 1 TFLOPS in computational capabilities. One Kepler GPU therefore provides more than 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared to the SIMD cores on CPUs, result in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that the actual throughput is highly algorithm dependent. Nevertheless, in theory, the computational power provided by modern GPUs is five times higher than that provided by CPUs. A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing incomparable hardware \cite{Lee2010, Kothapalli2013}. However, in the next section we will see why it is often easier to obtain an optimized implementation for the GPU than for the CPU.

@@

\begin{figure}[t!]
\centering
\subfloat[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
}
\subfloat[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
}
\caption{GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an in-place element-wise matrix square ((b) is from http://docs.nvidia.com/cuda).}
\label{fig:gpu_grid_kernel}
\end{figure}

@@

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction. Because of this the GPU is sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor. This also explain why GPU threads are said to be light-weighted. Such a conceptual difference means that there is an inherent focus on breaking algorithms down into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Manually typed SIMT, or SIMD, instructions is the only way of programming the GPU, while a single thread with automatic vectorization of for-loops is the standard way of programming the CPU.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something that is not likely to happen in practical applications, and therefore one will usually end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is an incomplete and minimal example of a GPU program. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is part of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, programming the GPU will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction. The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, GPU programming will hopefully be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

@@

\subsection{Beamforming}
If time delays are applied to the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images that are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$ planes. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
\caption{Delay-and-sum array beamforming. Image courtesy of Jo Inge Buskenes.}
\label{fig:das_background}
\end{figure}

@@

\subsection{Sampling and processing complexity}
When generating a B-mode image, we have to ensure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system that both emits and receives energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

@@

To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field. This is done by estimating a sample covariance matrix, which for an active broadband system can be done in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] &= \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H\\
\vec{x}_l[n'] &= [x_l[n'], \dotso, x_{l+L}[n']]
\end{align}
where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a subarray of length $L$, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.  Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}. The number of temporal samples is selected relative to the pulse length. To ensure robustness against model errors and numerical stability the matrix is further loaded with a diagonal factor, 
\begin{align}
\epsilon &= d\times\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align} 
This corresponds to a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.

@@

\section{Volume rendering}\label{sec:volren}
\begin{figure}[t!]
\centering
\subfloat[Ray-casting]{
	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
}
\subfloat[Opacity transfer function]{
	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
}
\caption{The principle behind volume ray casting. a) A ray is casted out from every pixel in the output image. Along the ray, samples are gathered from the volume. The samples are assigned a color and an opacity before they are combined into one pixel value (wikipedia.org/wiki/Volume\_ray\_casting). b) The opacity transfer function used to determine how opaque a sample should be.}
\label{fig:vr}
\end{figure}
A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user. One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers. Another way is to project the whole volume into a two-dimensional image. The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image. Many different methods for volume rendering exist \cite{brodlie2001}, but in this thesis we will focus on ray casting \cite{Levoy1988}.
